Leaf traits drive differences in biomass partitioning among major plant functional types

### 2015-07-23 manuscript.Rnw
Another revision. This one is very close, all paragraphs cleaned.

@@

The partitioning of biomass into leaves and stems is one of the most uncertain  and influential components of  global vegetation models (GVMs). Although GVMs typically assume that the major woody plant functional types (PFTs) differ in biomass partitioning, empirical data have been lacking. Using a new global database on biomass partitioning from over 175 studies, we here show that partitioning between stems and leaves varies strongly between PFTs (as defined by deciduous vs. evergreen and gymnosperm vs. angiosperm). We show that major woody PFTs do not differ in the amount of leaf area supported, but rather the amount of leaf mass per unit above ground biomass. Moreover, we find that these differences relate to systematic shifts in leaf mass per area, a leading trait within the leaf economics spectrum. In addition, we show how a second trait, leaf area per stem basal area, explains additional variation in biomass partitioning among species; however this trait does not differ systematically among major woody PFTs. Combined, these results uncover fundamental rules in how plants are constructed and allow for systematic benchmarking of biomass partitioning routines in current GVMs.

@@

Despite considerable advances in allometric theory (enquist etc. etc.) we don't have a clear picture of differences between PFTs. Instead, most of the work has focussed on understanding size-related shifts in biomass partitioning.  Longer leaf lifespan would logically lead to higher LMF (as argued by enquist, niklas) only if the annual allocation does not vary with leaf lifespan, but there is no a priori reason why this should be so. We take the view that a broader evidence base is needed to elucidate patterns which can subsequently be used to refine and improve existing models of allometry.

In this paper we are primarily interested in the distribution of biomass ('partitioning') between leaves and stems, an important component of the residence time of carbon in ecosystems \citep{friend2014}. Previous work based on whole stands \citep{oneilldeangelis1981,enquistniklas2002,reich2014} and individual plants \citep{poorter2012}, reported differences between angiosperms and gymnosperms in the amount of leaf biomass per unit above-ground biomass (the 'leaf mass fraction', LMF). \citet{poorter2012} and \citet{enquistniklas2002} attributed higher LMF in gymnosperms to longer leaf lifespan (LL) compared to typical angiosperms, but this begs the question whether differences in LMF are also apparent between deciduous and evergreen functional types within angiosperms. It is also unknown whether PFTs with higher leaf biomass also have higher total leaf area, which is relevant because it drives total light interception and thus productivity. Several classic studies have assumed, based only on differences in measurements of leaf mass, that gymnosperms carry more leaf area than angiosperms \citep{chabothicks1982, bond1989}.

Little is known about global-scale patterns in LMF and LAR in relation to climate. It can be expected that biomass partitioning is correlated with precipitation or mean annual temperature \citep{reich2014} because short-term experiments (cite) and climate gradients \citep{callaway1994,poyatos2007} have shown responsiveness of biomass partitioning to environmental drivers. A recent study \citep{reich2014} also demonstrated that stand-scale biomass partitioning was related to mean annual temperature across diverse forest stands. Indeed, many GVMs assume that climate affects biomass partitioning within a given PFT, often through soil water stress or other abiotic stress factors (see Notes S1). Again, however, these models do not agree on the degree of plasticity in biomass partitioning, or which variables should it respond to.

Here, we use a unique, new database \citep{Falster2015} (Fig. 1) to establish general rules on how biomass partitioning differs among three dominant woody PFTs across the globe: evergreen gymnosperms, evergreen angiosperms, and deciduous angiosperms. The dataset includes observations of biomass and size metrics for individual plants, contributed by 175 studies across nine vegetation types (Fig. S1). In this paper we include field-grown plants only, including the entire size range of woody plants (0.01 - \textgreater 100m height). One key challenge is thus to account for the very large size variation commonly found in any allometric variable \citep{niklas1994}. We do this by fitting a statistical model that does no assume a functional dependence on plant size, allowing us to study PFT patterns at a common plant height.

@@

We used generalized additive models (GAM) to visualize the relationships between biomass and plant size variables, and to estimate variables and their confidence intervals such as LMF at a common plant height. In all fitted GAMs, we used a cubic regression spline. For the smoothed term in the model (plant height), we used up to 3 or 4 degrees of freedom, which resulted in biologically realistic smoothed relationships. Within the GAM, we used a penalized regression smoother [@Wood2006] to allow the final degree of smoothness to be estimated from the data. In all fitted GAMs, we used species-dataset combination as a random effect.We used the \emph{mgcv} package in R \citep{wood2006}. All variables (except MAP and MAT) were log-transformed prior to analysis.

@@

We analyzed climate effects in two ways: with quantitative variables (mean annual precipitation, MAP; mean annual temperature, MAT), and by biome (tropical / temperate/ boreal), a simple classification taking into account both MAT and MAP (Fig. S1).

@@

Because of allometric relationships between the studied variables \citep{niklas1994}, there was a strong effect of plant height on LMF (Fig. 2a and S2). We corrected for plant height by estimating LMF at a common plant height, using a generalized additive model, a semi-parametric method that makes no assumption on the underlying functional relationships \citep{wood2006}. We then found that LMF at a common height was proportional to the average leaf mass per area (LMA) across the three PFTs (Fig. 2b). As a consequence, LAR was invariant between PFTs, because LAR = LMF / LMA (Fig. 2c and S3). Mirroring the results for LMF, we found that the amount of leaf mass per unit stem area ($M_F/A_S$) differs among the three PFTs, that these differences are also proportional to LMA (Fig. 2d), and that PFTs are similarly invariant in the amount of leaf area per unit stem area ($A_F/A_S$) (Fig. 2e).

@@

We found that decomposing as in Eq. \ref{eq:M1} was very useful to study climate effects because the second term absorbs nearly all of the size related variation in LMF (Tables \ref{table/varpartB} and S1) and is otherwise fairly conserved across PFTs (Fig. S4). As shown in Fig. 2b and Fig. 2d, the term $M_F/A_S$ exhibits comparable differences between PFTs to LMF, but unlike LMF, $M_F/A_S$ is nearly independent of plant height (Tables \ref{table/varpartB} and S1. $M_F/A_S$ is thus a useful proxy for LMF that can be compared across species and sites.

@@

We found that PFTs differed in LMF, and that LMF was proportional to LMA. This was true especially across the three PFTs studied, but also to a lesser extent across species within PFT. The consequence is that the amount of leaf area supported per unit biomass (leaf area ratio, LAR) did not differ between PFTs, and was much less variable within PFTs than LMF. This result is particularly convincing because we studied the entire size range of woody plants in natural forests (0.1-100m) (Fig. 2a). Previous studies have demonstrated a large difference in LMF between angiosperms and gymnosperms \citep{poorter2012}, but have not further separated functional types into deciduous vs. evergreen angiosperms, or provided an explanation in terms of leaf traits. We show that LMA -- a central trait of the leaf economics spectrum (LES)\citep{wrightetal2004} -- explains these differences in a consistent manner.

A previous large compilation of plant biomass partitioning \citep{poorter2012} (not restricted to field-grown plants) did not find a correlation between LMF and LMA across 8 PFTs (including non-woody PFTs). In their comparison, they did not correct for plant size, which has a very large effect on LMF (Fig. 2). It is certainly necessary to correct for plant size when comparing biomass partitioning parameters across PFTs.

The finding that LMF is proportional to LMA across PFTs implies that LAR does not differ. Why would this be the case? One explanation is that it is beneficial for plants to target a more constant LAR rather than LMF, because leaf area drives light interception. Because canopy size closely correlates with aboveground biomass \citep{duursma2010}, and the ratio of leaf area to canopy size determines self-shading and thus production efficiency of the foliage \citep{duursma20}, a low LAR is wasteful of woody biomass, and a high LAR wasteful of the foliage because self-shading becomes too high. Another, not mutually exclusive, explanation lies in the strong positive correlation between LMA and leaf lifespan (LL) \citep{wright2004}. Both \cite{poorter2012} and \citet{enquistniklas2002} explained higher LMF in gymnosperms by higher LL; plants simply maintain more foliage. However it is not not self-evident that else else is indeed equal, that is, annual production of foliage could differ between PFTs. Nonetheless, further supporting this explanation, \citet{reich1992} showed a positive correlation between LL and total stand leaf biomass across diverse forest stands.

We found that, as for LMF, $M_F/A_S$ was also proportional to LMA across PFTs (Fig. 2). Because LMF may be decomposed into the terms $M_F/A_S$ and $M_T/A_S$ (\ref{eq:M2}), this result suggests that $M_T/A_S$ does not vary between PFTs (see also Fig. S4). If this result holds, it means that aboveground biomass may be estimated across PFTs with a single equation. For tropical forests, \citet{chave2005} showed that aboveground biomass can be well estimated as,

@@

where F is a form factor, and $\rho$ wood density (kg m$^3$). In this equation leaf mass is subsumed in F, or assumed to be negligible (which hold for somewhat tall trees, see Fig. 2). Our result seems to indicate something like this works for all PFTs (see Fig. S4). However, there are differences in wood density between angiosperms and gymnosperms (Zanne + TRY categories), does this mean F is different between PFTs? It is timely to datamine BAAD to develop better general equations to estimate aboveground biomass.

Climate (temperature, rainfall) did not appreciably affect biomass partitioning between leaves and stems. This finding is in contrast with a recent study where LMF was found to increase with MAT across the globe \citep{reich2014}. One explanation for these differences is that our measurements were taken on individual plants, whereas in the other study they were on whole stands. This suggests climatic effects on stand-level biomass partitioning occur primarily by altering the size-distribution and stand density (number of individuals per unit area), rather than partitioning within individual plants. With the caveats that there is still variation between species and studies unaccounted for (Fig. S5), and that coarse climate variables such as MAT and MAP may mask small-scale climate effects on biomass partitioning, our results strongly suggest the effects of PFT are stronger than any direct climate effect on biomass partitioning within individual plants (Tables \ref{table/varpartB} and S1).

We found that PFTs differed in LMF, and that LMF was proportional to LMA. It thus seems possible to integrate a key leaf trait with whole-plant modelling of biomass partitioning. Indeed, nearly as much variation was explained across species when using LMA instead of PFT in a statistical model. This is surprising because LMA only captures aspects of leaf morphology, in indirectly leaf lifespan and photosynthetic capacity via the leaf economics spectrum \citep{wrightetal2004}. We thus show that in a PFT-based classification, LMA is a good first estimate of biomass partitioning, however, more variation can be explained by another trait ($A_F/A_S$), which varies appreciably between species. These results suggest promising avenues for parameterizing and simplifying biomass partitioning routines in GVMs.

>>>>

\renewcommand*{\thefootnote}{\fnsymbol{footnote}}

@@

The partitioning of biomass into leaves and stems is one of the most uncertain  and influential components of  global vegetation models (GVMs). Although GVMs typically assume that the major woody plant functional types (PFTs) differ in biomass partitioning, empirical studies have not been able to justify these differences. Using a new global database on biomass partitioning from over 175 studies, we here show that partitioning between stems and leaves varies strongly between PFTs (as defined by deciduous vs. evergreen and gymnosperm vs. angiosperm). We show that major woody PFTs do not differ in the amount of leaf area supported, but rather the amount of leaf mass per unit above ground biomass. Moreover, we find that these differences relate to systematic shifts in leaf mass per area, a leading trait within the leaf economics spectrum. In addition, we show how a second trait, leaf area per stem basal area, explains additional variation in biomass partitioning among species; however this trait does not differ systematically among major woody PFTs. Combined, these results uncover fundamental rules in how plants are constructed and allow for systematic benchmarking of biomass partitioning routines in current GVMs.

\emph{Keywords}: allocation, plant allometry, biomass estimation, specific leaf area, dynamic global vegetation model

@@

Despite considerable advances in theory underlying allometric scaling in plants \citep{enquist2003, west1999, savage2010} we do not yet have a clear understanding of potential differences between PFTs in terms of biomass partitioning. Instead, most previous work has focussed on understanding size-related shifts in biomass partitioning, as governed by constraints including hydraulic supply and mechanical stability. Rather than to advance a new model of biomass partitioning or allocation, we take the view that a broader evidence base is needed to elucidate patterns between PFTs. It is then up to the individual GVMs to refine the algorithms and parameter values to more closely match the observations.

In this paper we are primarily interested in the distribution of biomass ('partitioning') between leaves and stems, an important component of the residence time of carbon in ecosystems \citep{friend2014}. Previous work based on whole stands \citep{oneilldeangelis1981,enquistniklas2002,reich2014} and individual plants \citep{poorter2012}, reported differences between angiosperms and gymnosperms in the amount of leaf biomass per unit above-ground biomass (the 'leaf mass fraction', LMF). \citet{poorter2012} and \citet{enquistniklas2002} attributed higher LMF in gymnosperms to longer leaf lifespan (LL) compared to typical angiosperms, but this begs the question whether differences in LMF are also apparent between deciduous and evergreen functional types within angiosperms. It is also unknown whether PFTs with higher leaf biomass also have higher total leaf area, which is relevant because leaf area drives total light interception and thus productivity. Several classic, oft-cited, studies have assumed, based only on differences in measurements of leaf mass, that gymnosperms carry more leaf area than angiosperms \citep{chabothicks1982, bond1989}.  
Little is known about global-scale patterns in LMF and LAR in relation to climate. It can be expected that biomass partitioning is correlated with precipitation or mean annual temperature because experiments (cite) and comparisons across sites with contrasting climate \citep{callaway1994,poyatos2007} have shown responsiveness of biomass partitioning to environmental drivers. A recent study \citep{reich2014} demonstrated that stand-scale biomass partitioning was related to mean annual temperature across diverse forest stands. Indeed, many GVMs assume that climate affects biomass partitioning within a given PFT, often through soil water stress or other abiotic stress factors (see Notes S1). Again, however, these models do not agree on the degree of plasticity in biomass partitioning, or which climate variables it should respond to.

Here, we use a unique, new database \citep{Falster2015} (Fig. 1) to establish general rules on how biomass partitioning differs among three dominant woody PFTs across the globe: evergreen gymnosperms, evergreen angiosperms, and deciduous angiosperms. A second objective was to study relationships between biomass partitioning and climate variables or biome at a global scale. The dataset includes observations of biomass and size metrics for individual plants, contributed by 175 studies across nine vegetation types (Fig. S1), across the three major biomes (boreal, temperate and tropical). In this paper we include field-grown woody plants only, including the entire size range of woody plants (0.01 - \textgreater 100m height). One key challenge is thus to account for the very large size variation commonly found in any allometric variable \citep{niklas1994}. We do this by fitting a statistical model that does no assume a functional dependence on plant size, allowing us to study PFT patterns at a common plant height.

@@

We used generalized additive models (GAM) to visualize the relationships between biomass and plant size variables, and to estimate variables and their confidence intervals such as LMF at a common plant height. In all fitted GAMs, we used a cubic regression spline. For the smoothed term in the model (plant height), we used up to 3 or 4 degrees of freedom, which resulted in biologically realistic smoothed relationships. Within the GAM, we used a penalized regression smoother \citep{wood2006} to allow the final degree of smoothness to be estimated from the data. In all fitted GAMs, we used species-dataset combination as a random effect. We used the \emph{mgcv} package in R \citep{wood2006}. All variables (except MAP and MAT) were log-transformed prior to analysis.

@@

because we found that the vast majority of size-related variation was captured by $A_S/M_T$ alone. Similarly, we decomposed LAR as the product of $A_F/A_S$ and $A_S/M_T$. We analyzed climate effects on $M_F/A_S$ and $A_F/A_S$ in two ways: with quantitative variables (mean annual precipitation, MAP; mean annual temperature, MAT), and by biome (tropical / temperate/ boreal), a simple classification taking into account both MAT and MAP (Fig. S1). In both cases we used PFT and plant height as covariates.

@@

Because of allometric relationships between the studied variables \citep{niklas1994}, there was a strong effect of plant height on LMF (Fig. 2a and S2). When corrected for plant height by estimating LMF at a common plant height, we found that LMF was proportional to the average leaf mass per area (LMA) across the three PFTs (Fig. 2b). As a consequence, LAR was invariant between PFTs, because LAR = LMF / LMA (Fig. 2c and S3). Mirroring the results for LMF, we found that the amount of leaf mass per unit stem area ($M_F/A_S$) differs among the three PFTs, that these differences are also proportional to LMA (Fig. 2d), and that PFTs are similarly invariant in the amount of leaf area per unit stem area ($A_F/A_S$) (Fig. 2e).

@@

We found that decomposing LMF and LAR as shown in Eq. \ref{eq:M1} was very useful to study climate effects because the second term ($A_S/M_T$)absorbs nearly all of the size related variation in LMF (Tables \ref{table/varpartB}, Table S1 and Fig. S4) and is otherwise fairly conserved across PFTs (Fig. S4). As shown in Fig. 2b and Fig. 2d, the term $M_F/A_S$ exhibits comparable differences between PFTs to LMF, but unlike LMF, $M_F/A_S$ is nearly independent of plant height (Tables \ref{table/varpartB} and Table S1). $M_F/A_S$ is thus a useful proxy for LMF that can be compared across species and sites.

@@

We found that PFTs differed in LMF, and that LMF was proportional to LMA. This conclusion was particularly strong across the three PFTs studied, but it also held across species within PFT. The consequence is that the amount of leaf area supported per unit biomass (leaf area ratio, LAR) did not differ between PFTs. This result is particularly convincing because the data included individual plants from the entire size range of woody plants in natural forests (0.1-100m) (Fig. 2a). Previous studies have demonstrated a large difference in LMF between angiosperms and gymnosperms \citep{poorter2012}, but have not further separated functional types into deciduous vs. evergreen angiosperms, or provided an explanation in terms of leaf traits. We show that LMA -- a central trait of the leaf economics spectrum (LES) \citep{wrightetal2004} -- explains these differences in a consistent manner.

A previous large compilation of plant biomass partitioning \citep{poorter2012} (which included many non field-grown plants) did not find a correlation between LMF and LMA across 8 PFTs (including non-woody PFTs). However, they did not correct for plant size in their comparison. We found that, as expected, plant size strongly influences LMF (Fig. 2) and LAR (Fig. S3). It is certainly necessary to correct for plant size when comparing biomass partitioning parameters across PFTs\footnote{some better ending to this para is needed - not just want to bash poorter's crappy analysis, is there a positive note we can end on? we suggest a method to do so?}.

The finding that LMF is proportional to LMA across PFTs implies that LAR does not differ between PFTs. One explanation for this pattern lies in the strong positive correlation between LMA and leaf lifespan (LL) \citep{wright2004}. Both \cite{poorter2012} and \citet{enquistniklas2002} explained higher LMF in gymnosperms by higher LL; plants simply maintain more cohorts of foliage, but this explanation demands that all else is equal. It is not not self-evident that else else is indeed equal, that is, annual production of foliage could differ between PFTs. Nonetheless, further supporting this explanation, \citet{reich1992} showed a positive correlation between LL and total stand leaf biomass across diverse forest stands. Another, not mutually exclusive, possible explanation is that it may be beneficial for plants to target a more constant LAR rather than LMF, when considering the trade-off between leaf area and $M_T$. Canopy size scales with $M_T$ \citep{duursma2010}, and a larger canopy means higher light interception per unit leaf area \citep{duursma2007}. It can thus be argued that an optimal LAR exists that balances investment in supporting woody biomass (reflected by $M_T$) and the efficiency of foliage in terms of light interception.

We found that, as for LMF, $M_F/A_S$ was also proportional to LMA across PFTs (Fig. 2). Because LMF may be decomposed into the terms $M_F/A_S$ and $M_T/A_S$ (\ref{eq:M2}), this result suggests that $M_T/A_S$ does not vary between PFTs (see also Fig. S4). If this result holds, it means that aboveground biomass may be estimated across PFTs with an equation that does not differ between PFTs. For tropical forests, \citet{chave2005} showed that aboveground biomass can be well estimated as,

@@

where $F$ is a form factor, and $\rho$ wood density (kg m$^3$). Rearranging, this equation predicts that $M_T/A_S$ is proportional to plant height. We found some support for this prediction (Fig. S4). It is timely to use the BAAD to develop better, more general, equations to estimate individual plant aboveground biomass across the globe.

We found that climate did not appreciably affect biomass partitioning between leaves and stems. This finding is in contrast with a recent study where LMF was found to increase with MAT across the globe \citep{reich2014}. One explanation for this difference is that our measurements were taken on individual plants, whereas in  \citet{reich2014} they were on whole stands. This suggests climatic effects on stand-level biomass partitioning occur primarily by altering the size-distribution and stand density (number of individuals per unit area), rather than partitioning within individual plants. With the caveats that there is still variation between species and studies unaccounted for (Fig. S5), and that coarse climate variables such as MAT and MAP may mask small-scale climate effects on biomass partitioning, our results strongly suggest the effects of PFT are stronger than any direct climate effect on biomass partitioning within individual plants (Tables \ref{table/varpartB} and Table S1).

Overall, our results indicate that it is possible to integrate a key leaf trait with whole-plant modelling of biomass partitioning. Indeed, nearly as much variation was explained across species when using LMA instead of PFT in a statistical model. This is surprising because LMA only captures aspects of leaf morphology, in indirectly leaf lifespan and photosynthetic capacity via the leaf economics spectrum \citep{wrightetal2004}. We thus show that in a PFT-based classification, LMA is a good first estimate of biomass partitioning, however, more variation can be explained by another trait ($A_F/A_S$), which varies appreciably between species. These results suggest promising avenues for parameterizing and simplifying biomass partitioning routines in GVMs.

### 2015-07-23 manuscript_suppinfo.Rnw
Another revision. This one is very close, all paragraphs cleaned.

\usepackage{amssymb,amsmath}
\usepackage{palatino}
\usepackage{geometry}
\usepackage{authblk}
\usepackage{gensymb}
\geometry{verbose,tmargin=2cm,bmargin=2cm,lmargin=2cm,rmargin=2cm}
\graphicspath{ {./figures/} }
\usepackage{parskip}
\usepackage{natbib}

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.99\textwidth]{FigureS4.pdf}
    \caption{Relationship between above-ground biomass and basal stem area. Each point is an individual plant. Lines are spline fits. The right panel shows least-square means of the ratio of above-ground biomass to basal stem area, estimated from a linear mixed-effects model with basal stem area (linear and squared terms) and PFT. }
    \label{fig:figureS4}
\end{figure}

>>>>

\usepackage{amssymb,amsmath}
\usepackage{palatino}
\usepackage{geometry}
\usepackage{authblk}
\usepackage{gensymb}
\geometry{verbose,tmargin=2.4cm,bmargin=2.4cm,lmargin=2.4cm,rmargin=2.4cm}
\graphicspath{ {./figures/} }
\usepackage{parskip}
\usepackage{natbib}

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.99\textwidth]{FigureS4.pdf}
    \caption{Relationship between above-ground biomass and basal stem area. Each point is an individual plant. Lines are GAM fits by PFT. (a) Scaling of aboveground biomass ($M_T$) with basal stem area ($A_S$). (b) The ratio of $M_T/A_S$ as a function of plant height shows no clear differences between PFTs. The solid black line is a line with a slope of 1, which is expected if the aboveground biomass equation from \citet{chave2005} hold across the dataset. (c) Least-square means of the ratio of above-ground biomass to basal stem area, estimated from a linear mixed-effects model with basal stem area (linear and squared terms) and PFT. }
    \label{fig:figureS4}
\end{figure}
